
Red Wine Quality Exploration by Yiran Chen
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
#Packges
library(ggplot2)
library(gridExtra)
library(GGally)
library(corrplot)
library(psych)
library(wesanderson)
library(memisc)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
setwd("/home/niuniu/Documents/Udacity/P4/Assignment")
pf <- read.csv('wineQualityReds.csv')
```

This report explores a data set of 1599 red wines. Various exploration methods 
are used to analyse which chemical properties influence the quality of red 
wines. 

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots1}
str(pf)
summary(pf)
pf <- subset(pf, select = -c(X))
```

The data set contains 1599 red wines in total. There are thirteen variables. We 
are going to look at twelve variables, as variable "X" is line count. Hence 
variable "X" is removed from the dataset.


### Quality

```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots - Quality}
plot1 <- qplot(x = quality, data = pf, binwidth = 0.5, fill = I('#800020'))+
  scale_x_continuous(breaks = c(0:10))
plot1
summary(pf$quality)
table(pf$quality)
```

The best red wines among the data is rated eight, the worst receives a rating of
three. The graph shows that red wine quality is normally distributed with peaks 
at ratings five and six.


### Level of Alcohol

```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots - Alcohol}
summary(pf$alcohol)
qplot(alcohol, data = pf, fill = I('#800020'), binwidth = 0.25) +
  scale_x_continuous(breaks = seq(8,15,0.5))

```

The level of alcohol is skewed to the left. Mean of alcohol level is 10.43% and 
median is 9.5%.


### pH level

```{r echo=FALSE,message=FALSE, warning=FALSE, Univariate_Plots - pH}
summary(pf$pH)
qplot(pH, data = pf, fill = I('#800020'), binwidth = 0.1) +
  scale_x_continuous(breaks = seq(0,4.5, 0.1))

```

There are no basic wines and it is normally distributed with mean of 3.31 and 
median 3.310.


### Level of Acidity

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots - Acidity}
summary(pf$fixed.acidity)
qplot(fixed.acidity, data = pf, fill = I('#800020'), binwidth = 0.5)+
  scale_x_continuous(breaks = seq(0,16.5, 0.5))

summary(pf$volatile.acidity)
qplot(volatile.acidity, data = pf, fill = I('#800020'), binwidth = 0.05)+
  scale_x_continuous(breaks = seq(0,1.8, 0.1))

summary(pf$citric.acid)
qplot(citric.acid, data = pf, fill = I('#800020'), binwidth = 0.01) +
  scale_x_continuous(breaks = seq(0,1,0.05))

```

Both fixed acidity and volatile acidity are slightly skewed to the left. 
The plot for fixed acidity peaks around 7. There are few outliers within the 
volatile acidity plot. It peaks around 0.6 and the majority of the observations 
fall between 0.3 and 0.6. Additionally, all red wines contains a very small 
amount of citric acid and there are a peak at zero and another peak around 0.49.


### Total Sulfur dioxide

```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots - TotalSO2}

summary(pf$total.sulfur.dioxide)
qplot(total.sulfur.dioxide, data = pf, fill = I('#800020'), binwidth = 5) +
  scale_x_continuous(breaks = seq(0, 290, 10))

qplot(total.sulfur.dioxide, data = pf, fill = I('#800020'), binwidth = 0.05) +
  scale_x_log10(breaks = seq(0, 290, 20))


```

Total sulfur dioxide is skewed to the left and has a relatively long tail to the
right. In order to obtain a better understanding of the pattern, I have used 
log10 scale on x-axis. There is a peak around 45. The distribution is close to 
normal distribution.


### Free sulfur dioxide

```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots - FreeSO2}
summary(pf$free.sulfur.dioxide)
qplot(free.sulfur.dioxide, data = pf, fill = I('#800020'), binwidth = 2)+
  scale_x_continuous(breaks = seq(0, 75, 5))
  

qplot(free.sulfur.dioxide, data = pf, fill = I('#800020'), binwidth = 0.04)+
  scale_x_log10(breaks = seq(0, 72, 5))


```

Similar to total sulfur dioxide, the plot is skewed to the left. Logscale is 
used on x-axis. There is a peak around 6.

### Sulphates

```{r echo=FALSE, message=FALSE, warning=FALSE,Univariate_Plots - Sulphates}
summary(pf$sulphates)
qplot(sulphates, data = pf, fill = I('#800020'), binwidth = 0.1) +
  scale_x_continuous(breaks = seq(0,2, 0.1))
```

Sulphates also skew to the left and there are a few outliers which are greater 
than 1.5.


### Density of Wine

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots - Density}
summary(pf$density)
qplot(density, data = pf, fill = I('#800020')) +
  scale_x_continuous(breaks = seq(min(pf$density), max(pf$density), 0.002))


```

Density has a fairly small range overall. This is expected as wines have density
close to water. The distribution of red wine is normal with mean of 0.0067 
and median 0.9968.


### Level of residual sugar

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots - R_Sugar}
summary(pf$residual.sugar)
qplot(residual.sugar, data = pf, fill = I('#800020'))+
  scale_x_continuous(breaks = seq(min(pf$residual.sugar), 
                                  max(pf$residual.sugar), 1))
  

qplot(residual.sugar, data = pf, fill = I('#800020'))+
  scale_x_log10(breaks = seq(0, max(pf$residual.sugar), 1))
  

```

THe distribution is skewed to the left. The graph is transformed using a 
logscale on the x-axis. Majority of red wines have between around 1.6 g/dm^3 
to 2.8 g/dm^3. 

### Level of chlorides

```{r echo=FALSE, message=FALSE, warning=FALSE, Univariate_Plots - chlorides}
summary(pf$chlorides)
qplot(chlorides, data = pf, fill = I('#800020'))+
  scale_x_continuous(breaks = seq(min(pf$chlorides), max(pf$chlorides), 0.1))

qplot(chlorides, data = pf, fill = I('#800020'))+
  scale_x_continuous(breaks = seq(min(pf$chlorides), max(pf$chlorides), 0.1))+
  scale_x_log10(breaks = seq(0, max(pf$chlorides), 0.05))
  


```

Once again, chlorides plot is skewed to the left. I transformed the long tail 
data to better understand the distribution of chlorides. The distribution is 
close to normal if excluding observations above 0.2.


# Univariate Analysis

### What is the structure of your dataset?
There are 1599 observations within the data set. There 13 variables: fixed 
acidity, volatile acidity, citric acid, residual sugar, chlorides, free SO2, 
total SO2, density , pH, sulphates, alchohol, quality and X(excluded from 
analysis). All the variables are numerical.

### What is/are the main feature(s) of interest in your dataset?
The main feature is the quality in dataset. I would like to determine which 
chemical properties influence the quality of red wines. 


### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

I would like to include residual sugar, chlorides, the various types of acidity,
levels of different sulfur dioxides and density in the investigation. 

### Did you create any new variables from existing variables in the dataset?
No new variables created so far.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this? 

I noted a few variables, residual sugar, cholorides, total sulfur dioxide and 
free sulfur dioxide, have long tail data. Data transformation is performed using
a logscale on the x-axis. I was able to understand the pattern and distribution 
better. For instance, toal sulfur dioxide has a left skewed plot. After data 
transformation, it is noted that the distribution of total sulfur dioxide is 
close to normal distribution with a peak around 45.

# Bivariate Plots Section

In this section, I will find out the correlation between different variables, 
especially between quality and any other supporting variables. Then, I will pick
a few variables to perform further analysis.

### Overall correlation

```{r echo=FALSE,message=FALSE, warning=FALSE, Bivariate_Plots-corr1}
pairs.panels(pf, pch = 21, jiggle=TRUE)
```

The correlation plot above demonstrates correlation between variables. Due to 
the size of this dataset is fairly small, all the data are included in the plot.

It is noted that those variables have relatively strong correlation with red 
wine quality are alcohol (0.48) and volatile acidity (-0.39). I will investigate
these two variables further within this section. 

```{r echo=FALSE, message=FALSE, warning=FALSE,Bivariate_Plots-corr2}
M <- cor(pf)
plot2 <- corrplot(M, order="hclust", addrect = 2)
```


Similar to the plot above, this plot reveals the correlation between variables 
in a correlation matrix. The larger the circle, the stronger the correlation 
while blue means positive correlation and red indicates negative correlation. 
The rectangles around the chart of correlation matrix is drawn based on the 
results of hierarchical clustering.

Apart from the correlation between supporting variables and red wine quality, 
the following correlations are interesting: \
**alcohol vs. density** \
**fixed.acidity vs. density** \
**residual.sugar vs. density** \
**chlorides vs. sulphates** \
**free sulfur dioxide vs. total sulfur dioxide** \
**pH vs. fixed acidity** \
**pH vs. citric acid** \
\

The following relatively strong correlations are within expectation: \

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_FreeSO2&TotalSO2}
ggplot(data = pf, aes(y = free.sulfur.dioxide, x = total.sulfur.dioxide)) +
  geom_point()+
  geom_smooth(method = "lm")
```

**free sulfur dioxide vs. total sulfur dioxide** - These two variables have 
strong positive correlation of 0.67. The plot above illustrates positive 
correlation and the standard error around the trend line is relatively small. 
However, free sulfur dioxide is part of total sulfur dioxide. Hence it is within
norm that these two variables are correlated. \ 

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_PH&FA}
ggplot(data = pf, aes(y = pH, x = fixed.acidity)) +
  geom_point()+
  geom_smooth(method = "lm")
```

**pH vs. fixed acidity** - The above plot shows that fixed acidity and pH are 
negatively correlated ( -0.68) It is expected that an increase in fixed acidity 
in red wine will bring down the overall pH of the wine. Hence the negative 
correlation (- 0.68) between these two variables is expected. \


### Qualitiy and Alcohol

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Quality&Alcohol}
pf$quality_factor <- factor(pf$quality)
ggplot(data = pf, aes(y = alcohol, x = quality, color = quality_factor)) +
  geom_jitter(alpha = 1/4) + 
  geom_smooth(method = "lm")
boxplot1 <- ggplot(data = pf, aes(y = alcohol, x = quality_factor)) +
  geom_boxplot()
boxplot1
```

A new vaiiable, *quality_factor*, is created and this variable is a factor. I 
used "geom_jitter" to create the scatter plot to solve the issue of overlapping 
observations. From both scatter plot and box plot, it is noted that an increase 
in alcohol level is correlated to a higher quality of red wine. There is one 
exception, median alcohol level for rating 5
is smaller than median for rating 4. 

### Quality and Volatile Acidity

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Quality&V_Acidity}
ggplot(data = pf, aes(x = quality, y = volatile.acidity, 
                      color = quality_factor)) +
  geom_jitter(alpha =1/4)
boxplot2 <- ggplot(data = pf, aes(y = volatile.acidity, x = quality_factor)) +
  geom_boxplot()
boxplot2
```

From both scatter plot and box plot, it is noted that higher quality red wines 
is correlated with lower lever of volatile acidity. The boxplot shows a clearer 
pattern with the median volatile acidity decreasing as quality increases.


### Alcohol and Density

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Alcohol&Density}
ggplot(data = pf, aes(x = alcohol, y = density)) +
  geom_point()+
  geom_smooth(method = "lm")
```

A large proportion of observations have 9% to 11% alcohol. Alcohol and density 
are negatively correlated. 


### Fixed Acidity and Density

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_FixedAcidity&Density}
ggplot(data = pf, aes(x = fixed.acidity, y = density)) +
  geom_point()+
  geom_smooth(method = "lm")
```

The trend within the plot is clear, as fixed acidity increases, the density 
also increases.


### Residual Sugar and Density

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_sugar&density}
ggplot(data = subset(pf,residual.sugar <= quantile(pf$residual.sugar, 0.99)), 
       aes(x = residual.sugar, y = density)) +
  geom_point()+
  geom_smooth(method = "lm")
```

There are few outliners with large amount of residual sugar hence the top 1% 
population is excluded from the plot. Although there is an upward trend within 
the plot, the correlation between residual sugar and density is not very strong.
Majority of the observation have less than 4 g/dm^3. 


### Chlorides and Sulphates

```{r echo=FALSE, message=FALSE, warning=FALSE, Bivariate_Chlorides&Sulphates}
ggplot(data = subset(pf,chlorides <= quantile(pf$chlorides, 0.99)), 
       aes(x = chlorides, y = sulphates)) +
  geom_point()+
  geom_smooth(method = "lm")

ggplot(data = subset(pf,chlorides <0.15), aes(x = chlorides, y = sulphates)) +
  geom_point()+
  geom_smooth(method = "lm")

```

The top plot above exlcuded the top 1% of chlorides to remove outliers within 
the sample population. Similarly, there is an upward trend in the plot but the 
correlation between chlorides and sulphates does not appear to be strong. Most 
of the observations can be found to have between 0.05 g/dm^3 to 0.12 g/dm^3 of 
chlorides. 

In the second plot, only observations with chlorides below 0.15 g/dm^3. There 
does not appear to be a strong correlation between chlorides and sulphates.


# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features \
in the dataset?

In this section, I have plotted the correlation between variables using two 
different visualisations. I noted the two variables that have relatively strong 
correlation to red wine quality are alcohol and volatile acidity. It is noted 
that alcohol is positively correlated with red wine quality. Whereas an increse 
in volatile acidity leads to a decrease in red wine quality.

### Did you observe any interesting relationships between the other features

I also noted that alcohol and density is positively correlated. Similarly, fixed
acidity and density are also positively correclated.

### What was the strongest relationship you found?
The strongest relationship I found is between fixed acidity and pH which has a 
correlation of -0.68. 

# Multivariate Plots Section

### Red Wine Quality, Alcohol and Volatile Acidity

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_Plots}
plot3 <- ggplot(data = pf, aes(y = alcohol, x = volatile.acidity, 
                               color = quality_factor)) +
  geom_jitter(alpha = 1/4, size = 2)+
  scale_color_brewer(type = 'seq', palette = 'RdYlGn')
plot3
```

In the section above, I chose alcohol and volatile acidity to perform further 
analysis due to their relatively high correlation with red wine quality. In the 
plots above, all three variables are brought together. Color green represents 
good quality wines (7 and 8), yellow represents mid-range wines (5 and 6) and 
red for poor quality wines (3 and 4).For good quality wine, they appears to have
volatile acidity between 0.2 and 0.5 and alcohol between 11 and 13. 
Interestingly, wines with rating 5 sits at the bottom left of the plot. 
They have volatile acidity between 0.4 to 0.8 and alcohol level below 10. 

```{r echo=FALSE, message=FALSE, warning=FALSE, Multivariate_LM}
m1 <- lm(quality~alcohol + volatile.acidity, data = pf)
m2 <- update(m1, ~ . + density)
m3 <- update(m2, ~ . + fixed.acidity)
m4 <- update(m3, ~ . + residual.sugar)
m5 <- update(m4, ~ . + chlorides)
m6 <- update(m5, ~ . + sulphates)
mtable(m1, m2, m3, m4, m5, m6)
```

Based on plots in this section and the previous section, I created a linear 
model to included variables investigated in the past: quality, alcohol, valotile
acidity, density, fixed acidity, residual sugar, chlorides and sulphates. The 
variables in this linear model can account for only 30% of the variance in the 
quality of red wine. With the addition of variable, the R-squared value does not
improve. Hence the linear correlation between quality and the independent 
variables are weak.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

No. As the result of the linear regression model shows, there does not appear 
to be features that strengthen each other.

### Were there any interesting or surprising interactions between features?

From the plot, it is noted, good quality wines seems to have relatively high 
pecentage of alcohol and relatively low amount of volatile acidity.

### Did you create any models with your dataset? Discuss the strengths and \
limitations of your model.

A linear model was built for the dataset. It is a simple and easily 
interpretable model. However, due to the relatively weak linear correlation 
between variables, this might not be the best model to be used for this 
particular dataset.

------

# Final Plots and Summary


### Plot One
```{r echo=FALSE, Plot_One}
plot1 + 
  ggtitle('Red Wine Quality') + 
  theme(plot.title = element_text(hjust = 0.5))
```

### Description One

The distribution of red wine quality is normal and peaks at 5 and 6.

### Plot Two
```{r echo=FALSE, Plot_Two}
boxplot1.1 <- boxplot1 + 
  ggtitle('Quality By Alcohol (%)') + 
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab('alcohol (%)') +
  xlab('quality')

boxplot2.1 <- boxplot2 + 
  ggtitle('Quality By Volatile Acidity (g/dm^3)') + 
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab('volatile acidity (g/dm^3)')+
  xlab('quality')

grid.arrange(boxplot1.1, boxplot2.1, ncol =2)
```

### Description Two
Alcohol and volatile acidity are two variables that have relatively strong 
correlation with red wine quality. The two boxplots above show clear trends 
that quality and alcohol is positively correlated whereas volatile acidity is 
negatively correlated with red wine quality.


### Plot Three
```{r echo=FALSE, Plot_Three}
plot3 +
  ggtitle('Wine Quality By Alcohol (%) and Volatile Acidity (g/dm^3)') + 
  theme(plot.title = element_text(hjust = 0.5)) +
  ylab('alcohol (%)') +
  xlab('volatile acidity (g/dm^3)')
```

### Description Three
This plot has group all quality with both alcohol and volatile acidity. It is 
noted that higher quality wine (rating above 5) are correlated with higher 
percentage of alcohol and volatile acidity. Althought the overall trends shows 
red wine quality increase as percentage of alcohol increase, there is a dip at 
rating 5 where the median of alcohol is lower then rating 4. 

------

# Reflection

This data set contains information on 1,599 different red wine. The purpose of 
the analysis was to determine which chemical properties affect the quality of 
wine. I started my analysis by understanding each variable. Then, I explored the
correlation between each variables. None of the correlations were above 0.7. 
Among all variables, alcohol and volatile acidity had a relatively strong 
correlation with quality. Those two variables were picked for further analysis. 
Additionally, a few pairs of other variables were picked for analysis, such as 
density. Finally, both alcohol and volatile acidity were included in the plot 
with quality within multivariate analysis and a linear model was built. Per the 
result from the model, it is difficult to draw a definitive conclusion of which 
variables can affect the red wine quality significantly. \

One limitation of this dataset is that the sample size is relatively small. 
Although there are more almost 700 observations for rating 5 and 6, there are 
less than 70 observations under rating 3 and 4 and only 18 observations for 
rating 8. If there is any outlier within rating 3, 4 or 8, the overall trend 
will be inevitably affected. Another suggest will be include more categorical 
variables into the dataset, such production location which is an important 
factor or red wines.




